---
title: "How to create supercells"
author: "Givanna Putri"
package: SuperCellCyto
output: 
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{How to create supercells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, message=FALSE}
library(SuperCellCyto)
library(parallel)
```

# Introduction

This vignette describes the steps to reduce the size of vast high-dimensional cytometry data using SuperCellCyto, an R package based on the [SuperCell R package](https://github.com/GfellerLab/SuperCell) by David Gfeller lab from the University of Lausanne. 

Please note that we're still actively updating this vignette (and in fact the package itself),
and that we welcome any feedbacks on how to improve them.
There are myriad of ways on how to use SuperCell.
While we try to cover as many use cases as possible, we bound to miss something.
In that case, please reach out through the github repository by creating a Github issue.

# Installation

To install SuprCellCyto, we need to use the `devtools` package from CRAN. 
You can install `devtools` by using the `install.packages("devtools")` command.

Thereafter, you can install SuperCellCyto using 
`devtools::install_github("phipsonlab/SuperCellCyto")`.

SuperCellCyto requires the [SuperCell R package](https://github.com/GfellerLab/SuperCell)
installed to run properly.
If you use the `devtools::install_github` command above to install SuperCellCyto, 
it should be, in theory, automatically installed. 
But in the case it doesn't, you can manually install it by using
`devtools::install_github("GfellerLab/SuperCell")`.

# Preparing your dataset

The function which creates supercells is called `runSuperCellCyto`, and it 
operates on a `data.table` object, an enhanced version of R native `data.frame`.
We may add some support for `SummarizedExperiment` or `flowFrame` object
in the future if there are enough demands for it.

If the raw data is stored in a csv file, we can import it into a `data.table`
object using their `fread` function.

If the raw data is stored across multiple csv files or FCS files (more common for cytometry),
then we will need the help of [Spectre](https://github.com/ImmuneDynamics/Spectre) 
R package to import them as a`data.table` object.
Specifically, we need to:

1. Run `read.files` function to read in the FCS or csv files.
2. Run `do.merge.files` to merge the resulting `data.table` objects into one.

If you are unsure as to how these steps will work out, have a look at an example
in this [Spectre vignette](https://immunedynamics.io/spectre/simple-discovery/#2_Import_and_prep_data).

Using the vignette above, if you have csv files, you can run the steps in that vignette 
as they are, ***but*** only after changing the `InputDirectory` variable.
If you have FCS files, you need to change the `file.type` parameter for the 
`read.files` function to `.fcs`.

For this vignette, we will simulate some toy data using the `simCytoData` 
function.

```{r}
n_markers <- 15
n_samples <- 3
dat <- SuperCellCyto::simCytoData(nmarkers = n_markers, ncells = rep(10000, n_samples))
head(dat)
```

There are several things to note about our dataset. 
Let's go through them one by one in each sub-section below.

## The markers

The `runSuperCellCyto` function does not perform any data transformation or scaling.
Thus, we must ensure that our dataset have already been appropriately transformed 
using either the arc-sinh transformation or linear binning (using FlowJo).
This tutorial explains the data transformation process in very great detail:
(https://wiki.centenary.org.au/display/SPECTRE/Data+transformation).
Please have a read if you are unsure how to transform your data.

For our toy dataset, we will transform our data using the arc-sinh transformation 
implementation provided by the base R `asinh` function:

```{r}
# Specify which columns are the markers to transform
marker_cols <- paste0("Marker_", seq_len(n_markers))
# The co-factor for arc-sinh
cofactor <- 5

# Do the transformation
dat_asinh <- asinh(dat[, marker_cols, with = FALSE] / cofactor)

# Rename the new columns
marker_cols_asinh <- paste0(marker_cols, "_asinh")
names(dat_asinh) <- marker_cols_asinh

# Add them our previously loaded data
dat <- cbind(dat, dat_asinh)

head(dat[, marker_cols_asinh, with = FALSE])
```

Breaking down the steps, we:

1. Identify the columns denoting the markers.
2. Set the co-factor to 5.
3. Do the transformation and store it in `dat_asinh` variable.
4. Set the `dat_asinh` column name to reflect that the values in each column (marker) 
haas undergone an arc-sinh transformation.
5. Combine `dat` and `dat_asinh` using `cbind`.

## Cell id column

To create supercell, we must provide a column which uniquely identify each cell,
akin to the `Cell_Id` column in the toy data we generated above:

```{r}
head(dat$Cell_Id, n = 10)
```

The purpose of cell id is to allow SuperCell to uniquely identify each cell in the dataset.
This ID will come in super handy later when/if we need to work out which cells
belong to which supercells.

Generally, we will need to create this ID ourselves.
Most dataset won't come with this ID already embedded in.
A simple cell id can be made up by concatenating the word `Cell` with the row number.
Something like the following:

```{r}
dat$Cell_id_dummy <- paste0("Cell_", seq_len(nrow(dat)))
head(dat$Cell_id_dummy, n = 10)
```

Here, we store the cell id in a column called `Cell_id_dummy`.
It has values such as `Cell_1, Cell_2,` all the way until `Cell_x` where x is 
the number of cells in the dataset.

Note, we can name the cell id column however we like, `id, cell_identity`, etc. 
Importantly, we need make sure we note the column name as we will need to pass it 
to the `runSuperCellCyto` function later.

## Sample column

You will notice that in the toy data above, we have a column called `Sample`.
By default, this column refers to the biological sample the cells come from.
In the toy data above, we have 3 samples, `Sample_1, Sample_2, Sample_3`:

```{r}
unique(dat$Sample)
```

and we have 10,000 cells per sample:

```{r}
table(dat$Sample)
```

To create supercells, it is necessary to have this `Sample` column in our dataset.
We can name the column however we like, `Samp, Cell_Samp`.
However, we make sure we note the column name as we will need to pass it to the 
`runSuperCellCyto` function.
More on this in [Creating supercells][#creating-supercells] section. 

But what if we only have 1 biological sample in our dataset?
It does not matter.
We still need to have this column in our dataset, and pass the column name
to the `runSuperCellCyto` function.
The only difference is that this column will only have 1 unique value.

Why do we need to do this?
To ensure that each supercell only contains cells from exactly 1 sample.
This is because, in general, it does not make sense to mix cells from different 
biological samples in one supercell.
Additionally (not as important), the `runSuperCellCyto` function can process
all the samples in parallel if you supply its set the `n_parallel_worker` parameter
to greater than 1.
More on this in [Running runSuperCellCyto in parallel](#running-runsupercellcyto-in-parallel) 
section below.

However, if you want each supercell to contain cells from different biological samples,
then you need to create a new `Sample` column containing exactly 1 ***unique*** value, 
and pass the column name to `runSuperCellCyto` function.

\emph{Expert tips}:
You may wonder whether it is possible to use `SuperCellCyto` to reduce the number of cells
captured in each cluster (or cell type) so we can make a UMAP/tSNE plot that is not
as crowded?
Commonly in cytometry, we use stratified sampling to subsample our clusters before
drawing UMAP/tSNE plot to avoid overcrowding it. 

The short answer is, yes you can.
See [Using runSuperCellCyto for stratified summarising](#using-runsupercellcyto-for-stratified-summarising)
section for more information.

# Creating supercells

Now that we have imported our data, let's create some supercells.

First, let's store the markers, sample, and cell id column in variables:

```{r}
markers_col <- paste0("Marker_", seq_len(n_markers), "_asinh")
sample_col <- "Sample"
cell_id_col <- "Cell_Id_dummy"
```

Then pass all of that, together with the dataset into `runSuperCellCyto` function
to create supercells:

```{r}
supercells <- runSuperCellCyto(
  dt = dat,
  markers = markers_col,
  sample_colname = sample_col,
  cell_id_colname = cell_id_col
)
```

Now let's dig deeper into the object it created:

```{r}
class(supercells)
```

It is a list containing 3 elements:

```{r}
names(supercells)
```

## Supercell object

The `supercell_object` contains the metadata used to create the supercells.
It is a list, and each element contains the metadata used to create the supercells 
for a sample.
This will come in handy if we need to debug the supercells later down the line.

## Supercell expression matrix

The `supercell_expression_matrix` contains the marker expression of each supercell.
These are calculated by taking the average of the marker expression of all the cells 
contained within a supercell.

```{r}
head(supercells$supercell_expression_matrix)
```

Therein, we will have the following columns:

```{r}
names(supercells$supercell_expression_matrix)
```

1. All the markers we previously specified in the `markers_col` variable.
2. A column (`Sample` in this case) denoting which sample a supercell belongs to, 
(note the column name is the same as what is stored in `sample_col` variable).
3. The `SuperCellId` column denoting the unique ID of the supercell.

### SuperCellId
Let's have a look at `SuperCellId`:

```{r}
head(unique(supercells$supercell_expression_matrix$SuperCellId))
```

Let's break down one of them, `SuperCell_1_Sample_Sample_1`.
`SuperCell_1` is a numbering (1 to however many supercells there are in a sample)
used to uniquely identify each supercell in a sample.
Notably, you may encounter this (`SuperCell_1`, `SuperCell_2`) being repeated 
across different samples, e.g.,

```{r}
supercell_ids <- unique(supercells$supercell_expression_matrix$SuperCellId)
supercell_ids[grep("SuperCell_1_", supercell_ids)]
```

While these 3 supercells' id are pre-fixed with `SuperCell_1`, it does not make them
equal to one another! 
`SuperCell_1_Sample_Sample_1` will only contain cells from `Sample_1` while
`SuperCell_1_Sample_Sample_2` will only contain cells from `Sample_2`.

By now, you may have noticed that we appended the sample name into each
supercell id.
This aids in differentiating the supercells in different samples.

## Supercell cell map

`supercell_cell_map` maps each cell in our dataset to the supercell it belongs to.

```{r}
head(supercells$supercell_cell_map)
```

This map is very useful if we later need to expand the supercells out.
Additionally, this is also the reason why we need to have a column in the dataset 
which uniquely identify each cell.

# Running `runSuperCellCyto` in parallel

By default, `runSuperCellCyto` will process each sample one after the other. 
As each sample is processed independent of one another, we can process all of 
them in parallel. 
To do this, we need to set the `n_parallel_worker` to a number greater than 1.
Notably, we must not set more workers than the total number of cores we have in the computer.
To find out the total number of cores we have in the computer, we can use parallel's detectCores:
```{r}
n_cores <- detectCores()
```

We can thereafter supply this number as `n_parallel_worker` parameter.
Before we do that, do note that if you set `n_parallel_worker` to `n_core`, `runSuperCellCyto`
will utilise all your cores to process the samples, and render your computer rather
useless. 
So it is best to set this number to slightly less, something like `n_core - 1`:

```{r}
supercell_par <- runSuperCellCyto(
  dt = dat,
  markers = markers_col,
  sample_colname = sample_col,
  cell_id_colname = cell_id_col,
  n_parallel_worker = n_cores - 1
)
```


# Controlling the supercells' granularity

This is described in the `runSuperCellCyto` function's documentation, but let's 
briefly go through it here.

The `runSuperCellCyto` function is equipped with various parameters which can be
customise to alter the composition of the supercells.
The one is very likely to be used the most is the `gam` parameter.

The `gam` parameter controls how many supercells to generate, and 
indirectly, how many cells are captured within each supercell.
This parameter is resolved into the following formula `gam=n_cells/n_supercells`
where `n_cell` denotes the number of cells and `n_supercells` denotes the 
number of supercells. 

In general, the larger `gam` parameter is set to, the less supercells we will get.
Say for instance we have 10,000 cells. 
If `gam` is set to 10, we will end up with about 1,000 supercells, whereas
if `gam` is set to 50, we will end up with about 200 supercells.

You may have noticed, after reading the sections above, `runSuperCellCyto` is ran
on each sample independent of each other, and that we can only set 1 value as the `gam` parameter.
Indeed, for now, the same `gam` value will be used across all samples, and that depending
on how many cells we have in each sample, we will end up with different number
of supercells for each sample.
For instance, say we have 10,000 cells for sample 1, and 100,000 cells for sample 2.
If `gam` is set to 10, for sample 1, we will get 1,000 supercells (10,000/10) 
while for sample 2, we will get 10,000 supercells (100,000/10).

In the future, we may add the ability to specify different `gam` value for different
samples.
For now, if we want to do this, we will need to break down our data into multiple
`data.table` objects, each containing data from 1 sample, and run `runSuperCellCyto` 
function on each of them with different `gam` parameter value.
Something like the following:

```{r}
n_markers <- 10
dat <- simCytoData(nmarkers = n_markers)
markers_col <- paste0("Marker_", seq_len(n_markers))
sample_col <- "Sample"
cell_id_col <- "Cell_Id"

samples <- unique(dat[[sample_col]])
gam_values <- c(10, 20, 10)

supercells_diff_gam <- lapply(seq_len(length(samples)), function(i) {
  sample <- samples[i]
  gam <- gam_values[i]
  dat_samp <- dat[dat$Sample == sample, ]
  supercell_samp <- runSuperCellCyto(
    dt = dat_samp,
    markers = markers_col,
    sample_colname = sample_col,
    cell_id_colname = cell_id_col,
    gam = gam
  )
  return(supercell_samp)
})
```

Subsequently, to extract and combine the `supercell_expression_matrix` and 
`supercell_cell_map`, we will need to use `rbind`:

```{r}
supercell_expression_matrix <- do.call(
  "rbind", lapply(supercells_diff_gam, function(x) x[["supercell_expression_matrix"]])
)

supercell_cell_map <- do.call(
  "rbind", lapply(supercells_diff_gam, function(x) x[["supercell_cell_map"]])
)
```

```{r}
rbind(head(supercell_expression_matrix, n = 3), tail(supercell_expression_matrix, n = 3))
```

```{r}
rbind(head(supercell_cell_map, n = 3), tail(supercell_cell_map, n = 3))
```

# Using `runSuperCellCyto` for stratified summarising

As previously mentioned, we can use `runSuperCellCyto` to perform stratified summarising, 
i.e., to summarise (well, meaningfully sub-sample) each cluster or cell type.
To do this, we need to change the sample column such that it denotes
the cell type or the cluster a cell belongs to.

As an example, let's first cluster a toy data with k-means:

```{r}
set.seed(42)

# Simulate some data
dat <- simCytoData()
markers_col <- paste0("Marker_", seq_len(10))
cell_id_col <- "Cell_Id"

# Run kmeans
clust <- kmeans(
  x = dat[, markers_col, with = FALSE],
  centers = 5
)

clust_col <- "kmeans_clusters"
dat[[clust_col]] <- paste0("cluster_", clust$cluster)
```

To perform stratified summarising, we supply the cluster column
(`kmeans_clusters` in the example above), as `runSuperCellCyto`'s `sample_colname`
parameter.

```{r}
supercells <- runSuperCellCyto(
  dt = dat,
  markers = markers_col,
  sample_colname = clust_col,
  cell_id_colname = cell_id_col
)
```

Now, if we look at the `supercell_expression_matrix`, each row (each supercell) will
be denoted with the cluster it belongs to, and *not the biological sample it came from*:

```{r}
# Inspect the top 3 and bottom 3 of the expression matrix and some columns.
rbind(
  head(supercells$supercell_expression_matrix, n = 3),
  tail(supercells$supercell_expression_matrix, n = 3)
)[, c("kmeans_clusters", "SuperCellId", "Marker_10")]
```

If we look at the number of supercells created and check how many cells
there were in each cluster, we will find that, for each cluster, we get approximately
`n_cells/20` where 20 is the `gam` parameter value we used for `runSuperCellCyto`
(this is the default).

```{r}
# Compute how many cells per cluster, and divide by 20, the gamma value.
table(dat$kmeans_clusters) / 20
```

```{r}
table(supercells$supercell_expression_matrix$kmeans_clusters)
```

# Session information
```{r}
sessionInfo()
```
