---
title: "How to create supercells"
author: "Givanna Putri"
package: SuperCellCyto
output: 
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{How to create supercells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, message=FALSE}
library(SuperCellCyto)
```

# Introduction

This vignette describes the steps to reduce the size of vast high-dimensional cytometry data using SuperCellCyto, an R package based on the [SuperCell R package](https://github.com/GfellerLab/SuperCell) by David Gfeller lab from the University of Lausanne. 

Please note that we're still actively updating this vignette (and in fact the package itself),
and that we welcome any feedbacks on how to improve them.
There are myriad of ways on how to use SuperCell.
While we try to cover as many use cases as possible, we bound to miss something.
In that case, please reach out through the github repository.

# Installation

```{r, message=FALSE, warning=FALSE}
devtools::install_github("phipsonlab/SuperCellCyto")
```

SuperCellCyto requires the [SuperCell R package](https://github.com/GfellerLab/SuperCell)
installed to run properly.
If you use the installation command above, it should be, in theory, automatically installed. 
But in the case it doesn't, please manually install it by running the following command beforehand:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
devtools::install_github("GfellerLab/SuperCell")
```

# Preparing your dataset

The function which creates supercells is called `runSuperCellCyto`, and it 
operates on a `data.table` object, an enhanced version of R native `data.frame`.
We may add some support for `SummarizedExperiment` or `flowFrame` object
in the future if there are enough demands for it.

If the raw data is stored in a csv file, we can easily import it into a `data.table`
object using their `fread` function.

If the raw data is stored across multiple csv files or FCS files (more common for cytometry),
then we will need the help of [Spectre](https://github.com/ImmuneDynamics/Spectre) 
R package to import them as a`data.table` object.
Specifically, we need to:

1. Run `read.files` function to read in the FCS or csv files.
2. Run `do.merge.files` to merge the resulting `data.table` objects into one.

If you are unsure how these steps will work, have a look at this 
[Spectre vignette](https://immunedynamics.io/spectre/simple-discovery/#2_Import_and_prep_data).
If you have csv files, you can just run the steps in that vignette as they are
(of course after changing the `InputDirectory` variable).
If you have FCS files, just change the `file.type` parameter for `read.files`
function to `.fcs`.

For this vignette, we will simulate some data using the `simCytoData` 
function.

```{r}
n_markers <- 15
n_samples <- 3
dat <- SuperCellCyto::simCytoData(nmarkers = n_markers, nsample = n_samples)
head(dat)
```

There are several things to note about our dataset. 
Let's go through them one by one in each sub-section below.

## The markers

The `runSuperCellCyto` function does not perform any data transformation or scaling.
Thus, we must ensure that our dataset have already been appropriately transformed 
using either the arc-sinh transformation or linear binning (using FlowJo).
This tutorial explains the data transformation process in very great detail
(https://wiki.centenary.org.au/display/SPECTRE/Data+transformation).
Do have a read if you are unsure how to transform your data.

To perform arc-sinh transformation, we can simply use the native R `asinh` function:

```{r}
# Specify which columns are the markers to transform
marker_cols <- paste0("Marker_", seq_len(n_markers))
# The co-factor for arc-sinh
cofactor <- 5

# Do the transformation
dat_asinh <- asinh(dat[, marker_cols, with=FALSE] / cofactor)

# Rename the new columns
marker_cols_asinh <- paste0(marker_cols, "_asinh")
names(dat_asinh) <- marker_cols_asinh

# Add them our previously loaded data
dat <- cbind(dat, dat_asinh)

head(dat[, marker_cols_asinh, with=FALSE])
```

Breaking down the steps, we:

1. Identify the columns denoting the markers.
2. Set the co-factor to 5.
3. Do the transformation and store it in `dat_asinh` variable.
4. Set the `dat_asinh` column name to reflect that the columns have undergone arc-sinh transformation.
5. Combine `dat` and `dat_asinh` using `cbind`.


## Cell id column

To create supercell, we must provide a column which uniquely identify each cell,
akin to the `Cell_Id` column in the simulated data we just generated above:

```{r}
head(dat$Cell_Id, n=10)
```

This is to enable SuperCell to uniquely identify each cell in the dataset.
In fact, this ID will come in handy later if we need to work out which cells
belong to which supercells

Generally, we will need to create this ID ourselves. 
The good news is, it is very trivial to create it.
We can just number each cell from 1 to however many we have, and store that in 
a new column of our `data.table`. 
Something like the following:

```{r}
dat$Cell_id_dummy <- paste0("Cell_", seq_len(nrow(dat)))
head(dat$Cell_id_dummy, n=10)
```

Here, the column denoting the cell id is `Cell_id_dummy`, and it has values
such as `Cell_1, Cell_2,` all the way until `Cell_x` where x is the number of cells
there are in the dataset, each uniquely identifying a cell.

Our simulated dataset already comes with cell id attached to it (`Cell_Id`), yay!
So we will just use that from now. 

Note, we can name this column however we like, `id, cell_identity`, etc. 
Just make sure we note the column name as we will need to pass it into the 
`runSuperCellCyto` function later.

## Sample column

You will notice that in the simulated data above, we have a column called `Sample`.
By default, this column refers to the biological sample the cells come from.
In general, it is very common for a dataset to contain cells from multiple samples.

In the simulated data above, we have 3 samples, `Sample_1, Sample_2, Sample_3`:

```{r}
unique(dat$Sample)
```

and we have 10,000 cells per sample:

```{r}
table(dat$Sample)
```

To create supercells, it is necessary to have this `Sample` column in our dataset.
We can the column however we like, `Samp, Cell_Samp`.
Just make sure we note the column name as we need to pass it into the `runSuperCellCyto` function.
More on this in [Creating supercells][#creating-supercells] section. 

But what if we only have 1 biological sample in our dataset?
It does not matter.
We still need to have this column in our dataset, and pass the column name
into the `runSuperCellCyto` function.
The only difference is that this column will only have 1 unique value.

Why do we need to do this?
To ensure that each supercell only contain cells from exactly 1 sample.
This is because, in general, it does not make sense to mix cells from different 
biological samples in one supercell.
Additionally (not as important), the `runSuperCellCyto` function can process
all the samples in parallel if you supply its `BPPARAM` parameter with 
`BiocParallel::MulticoreParam`.
More on this in [Running runSuperCellCyto in parallel](#running-runsupercellcyto-in-parallel) 
section below.

However, if you want each supercell to contain cells from various biological samples,
then all you need to do is just treat your the cells as if they all come from one
biological sample, i.e. have only 1 unique value for the `Sample` column.

\emph{Expert tips}:
You may have noticed that `runSuperCellCyto` actually behaves a bit like 
stratified sampling, but instead of randomly sampling the cells from each population,
it summarises them.
In our example above, each population is a biological sample.

Hence, you may be thinking, can I then use `runSuperCellCyto` to perform stratified
summary of the clusters or cell types (rather than biological samples)?
In other words, if I have clustered or annotated my cells, can I use `runSuperCellCyto`
to reduce the size of each cluster or cell type in a meaningful way, rather than
randomly sub-sampling each cluster or cell type?

The short answer is, yes you can.
See [Using runSuperCellCyto for stratified summarising](#using-runsupercellcyto-for-stratified-summarising)
section for more information.

# Creating supercells

Now that we have imported our data, let's create some supercells.

First, let's store the markers, sample, and cell id column in variables:

```{r}
markers_col <- paste0("Marker_", seq_len(n_markers), "_asinh")
sample_col <- "Sample"
cell_id_col <- "Cell_Id"
```

Then pass all of that, together with the dataset into `runSuperCellCyto` function
to create supercells:

```{r}
supercells <- runSuperCellCyto(
    dt = dat,
    markers = markers_col,
    sample_colname = sample_col,
    cell_id_colname = cell_id_col
)
```

Now let's dig deeper into the object it created:

```{r}
class(supercells)
```

It is a list containing 3 elements:

```{r}
names(supercells)
```

## Supercell object

The `supercell_object` contains the metadata used to create the supercell objects.
It is a list, and each element contains the metadata used to create the supercells 
for a sample.
This will come in handy if we need to debug the supercells later down the line.

## Supercell expression matrix

The `supercell_expression_matrix` contains the marker expression of each supercell.
These are calculated by taking the average of the marker expression of all the cells 
contained within the supercell.

```{r}
head(supercells$supercell_expression_matrix)
```

Therein, we will have the following columns:

```{r}
names(supercells$supercell_expression_matrix)
```

1. All the markers we previously specified in `markers_col` variable.
2. A column (`Sample` in this case) denoting which sample a supercell belongs to, 
(note the column name is the same as what is stored in `sample_col` variable).
3. The `SuperCellId` column denoting the unique ID of the supercell.

### SuperCellId
Let's have a look at `SuperCellId`:

```{r}
head(unique(supercells$supercell_expression_matrix$SuperCellId))
```

Let's break down one of them, `SuperCell_1_Sample_Sample_1`.
`SuperCell_1` is a numbering (1 to however many supercells there are in a sample)
used to number each supercell in a sample.
Notably, you may encounter this being repeated across different samples, e.g.,

```{r}
supercell_ids <- unique(supercells$supercell_expression_matrix$SuperCellId)
supercell_ids[grep("SuperCell_1_", supercell_ids)]
```

While these 3 supercells' id are pre-fixed with `SuperCell_1`, it does not make them
equal to one another! 
`SuperCell_1_Sample_Sample_1` will only contain cells from `Sample_1` while
`SuperCell_1_Sample_Sample_2` will only contain cells from `Sample_2`.

By now, you would have noticed that we appended the sample name into each
supercell id.
This helps making each supercell id unique and more easily identifiable.

## Supercell cell map

`supercell_cell_map` maps each cell in our dataset to the supercell it belongs to.

```{r}
head(supercells$supercell_cell_map)
```

This map is very useful if we later need to revisit or re-analyse the cells contained
within a given supercell.
Additionally, this is also the reason why we need to have a column in the dataset 
which uniquely identify each cell.
Yes we could have just made a copy of the data and attach the supercell id 
as a new column, but this will result in larger RAM usage. 
Not ideal.

# Running `runSuperCellCyto` in parallel

By default, `runSuperCellCyto` will process each sample one after the other. 
As each sample is processed independent of one another, you can process all of 
them in parallel. 
To do this, you need to create a `BiocParallel::MulticoreParam` object and pass it
on as `BPPARAM` parameter:

```{r}
BPPARAM <- BiocParallel::MulticoreParam()
supercell_par <- runSuperCellCyto(
    dt = dat,
    markers = markers_col,
    sample_colname = sample_col,
    cell_id_colname = cell_id_col,
    BPPARAM = BPPARAM
)
```


# Controlling the supercells' granularity

This is described in the `runSuperCellCyto` function's documentation, but I'll 
repeat it here anyway. 

The `runSuperCellCyto` function is equipped with various parameters which you can
customise to change the composition of the supercells.
The one which we anticipate will be used the most is the `gam` parameter.

The `gam` parameter controls how many supercells to generate, and 
indirectly, how many cells are captured within a supercell.
This parameter is resolved into the following formula `gam=n_cells/n_supercells`
where `n_cell` denote the number of cells and `n_supercells` denote the number of
supercells. 

In general, the larger `gam` parameter is set to, the less supercells we will get.
Say for instance we have 10,000 cells. 
If `gam` is set to 10, we will end up with about 1,000 supercells, whereas
if `gam` is set to 50, we will end up with about 200 supercells.

You may have noticed, after reading the sections above, `runSuperCellCyto` is ran
on each sample independent of each other, and that we can only set 1 value as the `gam` parameter.
Indeed, for now, the same `gam` value will be used across all samples, and that depending
on how many cells we have in each sample, we will end up with different number
of supercells for each sample.
For instance, say we have 10,000 for sample 1, and 100,000 for sample 2.
If `gam` is set to 10, for sample 1, we will get 1,000 supercells (10,000/10) while for sample
2, we will get 10,000 supercells (100,000/10).

In the future, we may add the ability to specify different `gam` value for different
samples.
For now, if we want to do this, we will need to break down our data into multiple
`data.table` objects, each containing data from 1 sample, and run `runSuperCellCyto` 
function on each of them with differing `gam` parameter value.
Something like the following:

```{r}
n_markers <- 10
dat <- simCytoData(nmarkers = n_markers)
markers_col <- paste0("Marker_", seq_len(n_markers))
sample_col <- "Sample"
cell_id_col <- "Cell_Id"

samples <- unique(dat[[sample_col]])
gam_values <- c(10, 20, 10)

supercells_diff_gam <- lapply(seq_len(length(samples)), function(i) {
    sample <- samples[i]
    gam <- gam_values[i]
    dat_samp <- dat[dat$Sample == sample,]
    supercell_samp <- runSuperCellCyto(
        dt = dat_samp,
        markers = markers_col,
        sample_colname = sample_col,
        cell_id_colname = cell_id_col,
        gam = gam
    )
    return(supercell_samp)
})
```

To extract and combine the `supercell_expression_matrix` and `supercell_cell_map`,
we need to use `rbind`:

```{r}
supercell_expression_matrix <- do.call(
    "rbind",lapply(supercells_diff_gam, function(x) x[["supercell_expression_matrix"]])
)

supercell_cell_map <- do.call(
    "rbind",lapply(supercells_diff_gam, function(x) x[["supercell_cell_map"]])
)
```

```{r}
rbind(head(supercell_expression_matrix, n=3), tail(supercell_expression_matrix, n=3))
```

```{r}
rbind(head(supercell_cell_map, n=3), tail(supercell_cell_map, n=3))
```

# Using `runSuperCellCyto` for stratified summarising

As previously mentioned, we can use `runSuperCellCyto` to perform stratified summarising, 
i.e., to summarise (well, meaningfully subsample) each cluster or cell type.
To do this, all we need is to do is change the sample column such that it denotes
the cell type or cluster a cell belongs to.

As an example, let's cluster our simulated data with k-means:

```{r}
set.seed(42)

# Simulate some data
dat <- simCytoData()
markers_col <- paste0("Marker_", seq_len(10))
cell_id_col <- "Cell_Id"

# Run kmeans
clust <- kmeans(
    x = dat[, markers_col, with=FALSE],
    centers = 5
)

clust_col <- "kmeans_clusters"
dat[[clust_col]] <- paste0("cluster_", clust$cluster)
```

Now to do stratified summarising, all we need is supply the cluster column
(`kmeans_clusters` in the example above), as `runSuperCellCyto`'s `sample_colname`
parameter.

```{r}
supercells <- runSuperCellCyto(
    dt = dat,
    markers = markers_col,
    sample_colname = clust_col,
    cell_id_colname = cell_id_col
)
```

Now, if we look at the `supercell_expression_matrix`, each row (each supercell) will
be denoted with the cluster it belongs to, and *not the biological sample it came from*:

```{r}
# Let's show just the top 3 and last 3 of the expression matrix and some columns.
rbind(
    head(supercells$supercell_expression_matrix, n=3), 
    tail(supercells$supercell_expression_matrix, n=3)
)[, c('kmeans_clusters', 'SuperCellId', 'Marker_10')]
```

If we look at the number of supercells created and compare it with how many cells
there were in each cluster, we will find that, for each cluster, we get approximately
`n_cells/20` where 20 is the `gam` parameter value we used for `runSuperCellCyto`
(this is the default).

```{r}
# Compute how many cells per cluster, and divide by 20, the gamma value.
table(dat$kmeans_clusters) / 20
```

```{r}
table(supercells$supercell_expression_matrix$kmeans_clusters)
```

# Session information
```{r}
sessionInfo()
```



