---
title: "How to create supercells"
author: "Givanna Putri"
package: SuperCellCyto
output: 
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{How to create supercells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, message=FALSE}
library(SuperCellCyto)
```

# Introduction

This vignette describes the steps to meaningfully reduce the size of vast high-dimensional cytometry data using SuperCellCyto, an R package based on the [SuperCell R package](https://github.com/GfellerLab/SuperCell) by David Gfeller lab from the University of Lausanne. 

# Installation

```{r, message=FALSE, warning=FALSE}
remotes::install_github("phipsonlab/SuperCellCyto")
```

# Creating supercells

To create supercell representation of your data, you need to first store it as a `data.table` object, 
an enhanced version of R native `data.frame`.
For this example, we will just simulate some data using the `simCytoData()` function in this package.

If you have FCS files, you will need to somehow import them into `data.table`.
R package like [Spectre](https://github.com/ImmuneDynamics/Spectre) can help. 
Please consult them.

```{r}
n_markers <- 7
n_samples <- 3
dat <- SuperCellCyto::simCytoData(nmarkers = n_markers, nsample = n_samples)
head(dat)
```

A few extremely important things to note:

1. You ***must*** provide a column which uniquely identify each cell in *all your samples*.
This is to enable SuperCell to distinguish which cell belong to which sample and thus which supercell object.
See `Cell_id` column in the simulated data we just generated above.
2. All the samples must be concatenated into one `data.table` object, and there must a column in the `data.table` object which specifies which cell belong to which sample.
See `Sample` column in the simulated data we just generated above.
More on this below.

Explicitly specifying which cells belong to which sample ensures that a supercell does not group together cells from different samples.
If you think it is ok to mix cells from different samples in one supercell, 
then by all means just set the values in the `Sample` column to be the same for all cells.

If you have different treatment groups in your data, e.g., treated vs non-treated, 
it may not make sense for a supercell to contain cells belonging to both groups.
Say for instance you have 2 samples in each group: 
`treated_sample1`, `treated_sample2`, `non-treated_sample1`, `non-treated_sample2`.
For a supercell to not have cells from these samples mixed together,
you will need to specify, for each cell, which group and sample it belongs to as one value,
and store it in the `Sample` column. 
This can be easily achieved by concatenating the group and sample name with an underscore
or hyphen or whatever symbol (*not space character*) you like.

Note that I use `Cell_Id` and `Sample` columns in my example above.
You do not need to use the same column names.
Just make sure you note what they are.

In addition to the two columns above, also note the columns which denote the 
markers you want to use to create the supercells.
In most cases, it makes sense to include all the markers in your panel as it can help prevent different
cell types and states to be grouped together in a supercell. 

Let's put all of that together and create some supercells.

```{r}
markers <- paste0("Marker_", seq_len(n_markers))
supercells <- runSuperCellCyto(
    dt = dat,
    markers = markers,
    sample_colname = 'Sample',
    cell_id_colname = 'Cell_Id'
)
names(supercells)
```

The supercell object should be a list containing 3 elements: `supercell_object`,
`supercell_expression_matrix`, and `supercell_cell_map`.
The `supercell_object` contains the metadata used to create the supercell objects.

`supercell_expression_matrix` gives you the marker expression of each of the supercell.
These are calculated by taking the average of the marker expression of all the cells captured by the supercell.

```{r}
head(supercells$supercell_expression_matrix)
```

`supercell_cell_map` maps each cell to the supercell it belongs to.

```{r}
head(supercells$supercell_cell_map)
```

Depending on what you want to use supercell for, you would want to use different
element of the result list. 

Generally, your aim would be to reduce your data for subsequent downstream analysis.
In this case, run all your analysis steps on the `supercell_expression_matrix`.
Afterwards, if you want to drill into few individual cells (say you have clustered
the supercells, and you want to just drill into the cells captured within the 
supercells belonging to a given cluster), then you can use the `supercell_cell_map`
to obtain which cells were captured in which supercells, isolate them, and drill into them.

# Session information
```{r}
sessionInfo()
```



