---
title: "Interoperability with Seurat"
author: "Givanna Putri"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{interoperability_with_seurat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

How can we integrate SuperCellCyto's output with cytometry data stored in [Seurat](https://satijalab.org/seurat/) objects?
Is it possible to store the results of SuperCellCyto as Seurat objects for subsequent analysis using Seurat?
The answer to both is yes.

In this vignette, we demonstrate this process using a subsampled Levine_32dim dataset
used in our manuscript stored in a Seurat object.
We will show how to create supercells from this data and analyze them using [Seurat](https://satijalab.org/seurat/).

```{r setup, message=FALSE, warning=FALSE}
library(qs)
library(Seurat)
library(data.table)
library(SuperCellCyto)

set.seed(42)
```

## Preparing Seurat object

The subsampled Levine_32dim data is stored as a [qs](https://cran.r-project.org/web/packages/qs/index.html) object.
We begin by loading this data using the `qread` function from the qs package.


```{r}
seurat_obj <- qread("data/Levine_32dim_seurat_sub.qs")
seurat_obj
```
The data is stored in the `originalexp` assay, with both counts and data slots containing raw data.

Before running SuperCellCyto, we will first:

1. Subset this data to retain only the markers we need to perform downstream analysis.
2. Perform arcsinh transformation, and store the transformed data in the `data` slot of the `originalexp` assay.

```{r}
markers <- c(
    "CD45RA", "CD133", "CD19", "CD22", "CD11b", "CD4", "CD8", "CD34", "Flt3",
    "CD20", "CXCR4", "CD235ab", "CD45", "CD123", "CD321", "CD14", "CD33", "CD47",
    "CD11c", "CD7", "CD15", "CD16", "CD44", "CD38", "CD13", "CD3", "CD61", "CD117",
    "CD49d", "HLA-DR", "CD64", "CD41"
)

# keep only the relevant markers
seurat_obj <- seurat_obj[markers, ]

# to store arcsinh transformed data
seurat_obj[['originalexp']]$data <- asinh(seurat_obj[['originalexp']]$counts / 5)

seurat_obj
```

## Run SuperCellCyto

SuperCellCyto requires the input data in a `data.table` format.
Hence, we will extract the arcsinh-transformed data from the Seurat object,
format it into a data.table, and include sample information and cell IDs.

It's important to note that Seurat objects typically store cells as columns and features (markers or genes) as rows.
Since SuperCellCyto expects cells as rows and features as columns, we will also transpose
the data.

After transposing and preparing the `data table`, we run `runSuperCellCyto` function,
passing the required parameters including the markers, sample column name, cell ID column name, and gamma value.

```{r}
# check.names set to FALSE so HLA-DR is not replaced with HLA.DR
dt <- data.table(t(data.frame(seurat_obj[['originalexp']]$data, check.names = FALSE)))
# add the cell_id and sample metadata
dt <- cbind(dt, seurat_obj[[c("cell_id", "sample")]])

supercells <- runSuperCellCyto(
    dt = dt,
    markers = markers,
    sample_colname = "sample",
    cell_id_colname = "cell_id",
    gam = 5
)

head(supercells$supercell_expression_matrix)
```

We can now embed the supercell ID in the metadata of our Seurat object.

```{r}
seurat_obj$supercell_id <- factor(supercells$supercell_cell_map$SuperCellID)
head(seurat_obj[[]])
```

## Analyse Supercells as Seurat object

The supercell expression matrix, having fewer supercells than the number of cells in the Seurat object,
is best stored as a separate Seurat object.
This allows us to use Seurat's functions for analysis.

To do this, we first transpose the expression matrix, ensuring cells are columns and markers are rows,
and then create a new Seurat object with the default `RNA` assay.
The `data` and `counts` slots of the RNA assay are then set to contain the marker expression.

```{r}
supercell_exp_mat <- t(supercells$supercell_expression_matrix[, markers, with=FALSE])
colnames(supercell_exp_mat) <- supercells$supercell_expression_matrix$SuperCellId
supercell_seurat_obj <- CreateSeuratObject(counts = supercell_exp_mat)
supercell_seurat_obj[['RNA']]$data <- supercell_seurat_obj[['RNA']]$counts

supercell_seurat_obj
```
With the supercell marker expression stored as a Seurat object, we can proceed with
performing downstream analysis such as clustering and creating UMAP plots.

```{r message=FALSE}
# Have to do this, otherwise Seurat will complain
supercell_seurat_obj <- ScaleData(supercell_seurat_obj)

supercell_seurat_obj <- RunPCA(
    object = supercell_seurat_obj,
    npcs = 10,
    nfeatures.print = 1,
    approx = FALSE,
    seed.use = 42,
    features = markers
)

supercell_seurat_obj <- FindNeighbors(supercell_seurat_obj, dims = 1:10)
supercell_seurat_obj <- FindClusters(supercell_seurat_obj, resolution = 0.5)
supercell_seurat_obj <- RunUMAP(supercell_seurat_obj, dims = 1:10)

DimPlot(supercell_seurat_obj, reduction = "umap")
```

```{r height=10, width=10}
FeaturePlot(supercell_seurat_obj, features = c("CD4", "CD8", "CD19", "CD34", "CD11b"), ncol = 3)
```

## Transfer information from supercells Seurat object to single cell Seurat object

To transfer information (e.g., clusters) obtained from analyzing supercells back to the single cells,
we need to do some data wrangling.
The key is ensuring the order of the new information aligns with the order of cells in the Seurat object.

We demonstrate this using cluster information as an example.
We first extract the metadata from the single-cell Seurat object and clustering information
from the supercells Seurat object into two different `data.table` objects,
We then merge them using `merge.data.table`, setting the `sort` parameter to FALSE
and `x` parameter to the `data.table` containing the metadata from the single-cell Seurat object
These ensure the result is in the order of the metadata from our single-cell Seurat object.

```{r}
clusters <- data.table(
    supercell_id = colnames(supercell_seurat_obj),
    cluster = as.vector(Idents(supercell_seurat_obj))
)

cell_metadata <- seurat_obj[[]]
cell_metadata <- merge.data.table(
    x = cell_metadata,
    y = clusters,
    by = "supercell_id",
    sort = FALSE
)
```

After merging, we can add the cluster assignment to the metadata of the single-cell Seurat object.

```{r}
seurat_obj$cluster <- cell_metadata$cluster
Idents(seurat_obj) <- "cluster"
```

Then visualise the cluster assignments and marker expressions of our clustered single cell data.

```{r message=FALSE}
seurat_obj <- ScaleData(seurat_obj)
seurat_obj <- RunPCA(
    object = seurat_obj,
    npcs = 10,
    nfeatures.print = 1,
    approx = FALSE,
    seed.use = 42,
    features = markers
)

seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10)
seurat_obj <- FindClusters(seurat_obj, resolution = 0.5)
seurat_obj <- RunUMAP(seurat_obj, dims = 1:10)

DimPlot(seurat_obj, reduction = "umap")

```

```{r}
FeaturePlot(seurat_obj, features = c("CD4", "CD8", "CD19", "CD34", "CD11b"), ncol = 3)
```


## Session information

```{r}
sessionInfo()
```