---
title: "Using SuperCellCyto for Stratified Summarising"
author: "Givanna Putri"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-runsupercellcyto-for-stratified-summarising}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Have you been following the vignette on [how to create supercells](how_to_create_supercells.html),
and wonder whether it is possible to use `SuperCellCyto` as a replacement for stratified sampling
to avoid overcrowding UMAP/tSNE plot?

The short answer to that is, yes we can.
We call this **stratified summarising**, and `SuperCellCyto` can absolutely be used for this purpose.
To do this, all we need to do is simply set the sample column of our data to not be the
biological sample the cell came from, but rather` the column the we want to stratify the data based on.

For example, when drawing UMAP or tSNE plot, we commonly subsample each cluster or cell type to avoid
crowding the plot.
Instead of subsampling, we can generate supercells for each cluster or cell type simply
by specifying the column that denotes the cluster or cell type each cell belong to
as the `sample_colname` parameter!

Let's illustrate this using a clustered (using k-means) toy data.

```{r}
library(SuperCellCyto)

set.seed(42)

# Simulate some data
dat <- simCytoData()
markers_col <- paste0("Marker_", seq_len(10))
cell_id_col <- "Cell_Id"

# Run kmeans
clust <- kmeans(
  x = dat[, markers_col, with = FALSE],
  centers = 5
)

clust_col <- "kmeans_clusters"
dat[[clust_col]] <- paste0("cluster_", clust$cluster)
```

To perform stratified summarising, we supply the cluster column
(`kmeans_clusters` in the example above), as `runSuperCellCyto`'s `sample_colname`
parameter.

```{r}
supercells <- runSuperCellCyto(
  dt = dat,
  markers = markers_col,
  sample_colname = clust_col,
  cell_id_colname = cell_id_col
)
```

Now, if we look at the `supercell_expression_matrix`, each row (each supercell) will
be denoted with the cluster it belongs to, and *not the biological sample it came from*:

```{r}
# Inspect the top 3 and bottom 3 of the expression matrix and some columns.
rbind(
  head(supercells$supercell_expression_matrix, n = 3),
  tail(supercells$supercell_expression_matrix, n = 3)
)[, c("kmeans_clusters", "SuperCellId", "Marker_10")]
```

If we look at the number of supercells created and check how many cells
there were in each cluster, we will find that, for each cluster, we get approximately
`n_cells_in_the_cluster/20` where 20 is the `gam` parameter value we used for `runSuperCellCyto`
(this is the default).

```{r}
# Compute how many cells per cluster, and divide by 20, the gamma value.
table(dat$kmeans_clusters) / 20
```

```{r}
table(supercells$supercell_expression_matrix$kmeans_clusters)
```

## Session information
```{r}
sessionInfo()
```



